A Machine Profile Summary

### 2022-06-13 MachineSummary/MachineSummary.Rmd
updated machine summary

---
title: "A Machine Profile Summary"
author: 
   - John C. Nash \thanks{ retired professor, Telfer School of Management, University of Ottawa}
   - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
date: "14/06/2021 revised 18/01/2022"
bibliography: ../BibSupport/ImproveNLS.bib
output: 
  pdf_document:
    keep_tex: false
---

@@

```{r tdstamp}
#get a timestamp
tsstr<- format(Sys.time(), "%Y%m%d%H%M") # tsstr == time stamp string in form YYYYmmddHHMM
cat("Date and Time stamp:",tsstr,"\n")
```

@@

An important tool for acquiring machine information is the CRAN package
`benchmarkme` (@benchmarkme). We found a minor bug in the `get_ram()`
function of this package for some Windows 10 versions which is now 
corrected (??check it is uploaded). ??Do we wish to comment on the
"GB" issue?

@@

```{r benchmarkmex}
library(benchmarkmeData)
library(benchmarkme)
ls(package:benchmarkmeData)
ls(package:benchmarkme)
lsf.str("package:benchmarkme")
lsf.str("package:benchmarkmeData")
## This next line takes a lot of time to run, so is commented out here
# benchmark_std()
get_byte_compiler()
gla<-get_linear_algebra()
gsd<-get_sys_details()
```

@@

Until the IEEE 754 standard (??ref) became availaible in the 1980s, 
computers used multiple forms of floating point arithmetic. With
some ingenuity, it is possible to detect many aspects of the 
floating point properties.
This is a port of Mike Malcolm's ENVRON to R. ??ref. It computes the 
machine precision, radix and number of radix digits.

@@

```{r envrontest}
envron<-function(){
D1<-as.numeric(1) # use variable to avoid constants when double precision invoked
E5 <- 10 #arbitrary scaling for additive equality tests
B9 <- 1E+35 # big number, not necessarily biggest possible
E6 <- 1 # initial value for radix 
E9 <- 1 # initial value for machine precision
D0 <- E6 
repeat{
  E9 <- E9/2 #start of loop to decrease estimated machine precision
  D0 <- E6+E9 #force storage of sum into a floating-point scalar
  if (D0 <= E6) break
} # repeat reduction while (1+E9) > 1
E9 <- E9*2 # restore smallest E9 which gives (1+E9) > 1
repeat {
   E6 <- E6+1 # try different radix values
   D0 <- E6+E9
   if (D0 <= E6) break
} # until a shift is observed
J1 <- 1 # initial count of radix digits in mantissa
E9 <- 1 # use radix for exact machine precision
repeat {  
  E9 <- E9/E6 #loop while dividing by radix
  J1 <- J1+1 #increment counter
  D0 <- D1+E9 #add tp 1
  if (D0<=D1) break #test and repeat until equality
}
E9 <- E9*E6 #recover last value of machine precision
J1 <- J1-1 #and adjust the number of digits
mpvals<-list(eps=E9, radix=E6, ndigits=J1)
}
mp<-envron()
E9<-mp$eps
E6<-mp$radix
J1<-mp$ndigits
cat("E9 = ",E9,"  -- the machine precision,  E9=MIN (X  1+X>1)\n")
cat( "E9*1E+16=",E9*1E+16,"\n")
cat( "E6 = ",E6,"  -- the radix of arithmetic","\n")
cat( "J1 = ",J1,"  -- the number of radix digits in mantissa of FP numbers","\n")
cat( "E6^(-J1+1) * 1E+16=",1E+16*(E6^(-J1+1)),"\n")
cat(".Machine$double.eps=",.Machine$double.eps,"\n")
cat("From R  .Machine:\n")
cat("double.eps=", .Machine$double.eps,"\n", 
    "double.base=", .Machine$double.base,"\n", 
    "double.digits=", .Machine$double.digits,"\n" )
```

@@

```{r loopesc}
loopesc <- function(nn){
  ss<-0
  for (i in 1:nn) {
    xx<-exp(sin(cos(1.0*i)))
    ss<-ss+xx
  }
  xx
}
require("microbenchmark")
nvals<-c(1000, 10000, 100000, 1000000)
tvals<-rep(NA,length(nvals))
i<-0
for (nn in nvals) {
  i<-i+1
  cat(nn,"\n")
  tt<-microbenchmark(loopesc(nn), unit='us')
  print(tt)
  tvals[i]<-mean(tt$time)
}
plot(log10(nvals),log(tvals))

>>>>

---
title: "A Machine Profile Summary"
author: 
   - John C. Nash \thanks{ retired professor, Telfer School of Management, University of Ottawa}
   - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
date: "14/06/2021 revised 18/01/2022"
bibliography: ../BibSupport/ImproveNLS.bib
output: 
  pdf_document:
    keep_tex: false
link-citations: yes
linkcolor: blue
---

@@

```{r tdstamp}
#get a timestamp
tsstr <- format(Sys.time(), "%Y/%m/%d|%H:%M") # tsstr == time stamp string in form YYYYmmddHHMM
cat("Date and Time stamp:", tsstr, "\n")
```

@@

An important tool for acquiring machine information is the CRAN package
`benchmarkme` (@benchmarkme). We found a minor bug in the `get_ram()`
function of this package for some Windows 10 versions. It has been 
corrected in the latest version 1.0.8, with contributions from one of the authors (AB).

<!-- on https://github.com/csgillespie/benchmarkme, but at the -->
<!-- time of writing is not yet on the CRAN repository. ??Do we wish to comment on the -->
<!-- "GB" issue? -->

@@

```{r benchmarkmex}
library(benchmarkmeData)
library(benchmarkme)
ls(package:benchmarkmeData)
ls(package:benchmarkme)
lsf.str("package:benchmarkme")
lsf.str("package:benchmarkmeData")
## This next line takes a lot of time to run, so is commented out here
# benchmark_std()
get_byte_compiler()
gla <- get_linear_algebra()
gsd <- get_sys_details()
```

@@

Until the IEEE 754 standard (?? @institute1985ieee) became available in the 1980s, 
computers used multiple forms of floating point arithmetic. With
some ingenuity, it is possible to detect many aspects of the 
floating point properties.
This is a port(??part) of Mike Malcolm's ENVRON to R. ??ref. It computes the 
machine precision, radix and number of radix digits.

@@

```{r envrontest}
envron <- function(){
D1 <- as.numeric(1) # use variable to avoid constants when double precision invoked
E5 <- 10 # arbitrary scaling for additive equality tests
B9 <- 1E+35 # big number, not necessarily biggest possible
E6 <- 1 # initial value for radix 
E9 <- 1 # initial value for machine precision
D0 <- E6 
repeat{
  E9 <- E9 / 2 # start of loop to decrease estimated machine precision
  D0 <- E6 + E9 # force storage of sum into a floating-point scalar
  if (D0 <= E6) break
} # repeat reduction while (1+E9) > 1
E9 <- E9 * 2 # restore smallest E9 which gives (1 + E9) > 1
repeat {
   E6 <- E6 + 1 # try different radix values
   D0 <- E6 + E9
   if (D0 <= E6) break
} # until a shift is observed
J1 <- 1 # initial count of radix digits in mantissa
E9 <- 1 # use radix for exact machine precision
repeat {  
  E9 <- E9 / E6 # loop while dividing by radix
  J1 <- J1 + 1 # increment counter
  D0 <- D1 + E9 # add tp 1
  if (D0 <= D1) break # test and repeat until equality
}
E9 <- E9 * E6 # recover last value of machine precision
J1 <- J1 - 1 # and adjust the number of digits
mpvals<-list(eps = E9, radix = E6, ndigits = J1)
}
mp <- envron()
E9 <- mp$eps
E6 <- mp$radix
J1 <- mp$ndigits
cat("E9 = ",E9,"  -- the machine precision,  E9=MIN (X  1+X>1)\n")
cat( "E9*1E+16=", E9 * 1E+16, "\n")
cat( "E6 = ",E6,"  -- the radix of arithmetic", "\n")
cat( "J1 = ",J1,"  -- the number of radix digits in mantissa of FP numbers", "\n")
cat( "E6^(-J1+1) * 1E+16=", 1E+16*(E6^(-J1+1)), "\n")
cat(".Machine$double.eps=", .Machine$double.eps, "\n")
cat("From R  .Machine:\n")
cat("double.eps =", .Machine$double.eps, "\n", 
    "double.base =", .Machine$double.base, "\n", 
    "double.digits =", .Machine$double.digits, "\n" )
```

@@

```{r loopesc}
loopesc <- function(nn){
  ss <- 0
  for (i in 1:nn) {
    xx <- exp(sin(cos(1.0*i)))
    ss <- ss + xx
  }
  xx
}
require("microbenchmark")
nvals<-c(1000, 10000, 100000, 1000000)
tvals<-rep(NA, length(nvals))
i<-0
for (nn in nvals) {
  i <- i + 1
  cat(nn, "\n")
  tt <- microbenchmark(loopesc(nn), unit = 'us')
  print(tt)
  tvals[i] <- mean(tt$time)
}
plot(log10(nvals), log(tvals))
